\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-05-05 Writeup/kohler_clarke_2021.tex
most of my answers to R1

\documentclass[11pt, twoside]{article}
\usepackage{boisik}
\usepackage[OT1]{fontenc}
\usepackage[affil-it]{authblk}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{amsmath}
 \geometry{a4paper, total={170mm,257mm}, left=20mm, top=20mm}

@@

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \citep{RN1425}. Wallpaper groups differ in the number and kind of these transformations. In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \citep{RN1425}. The full list of subgroup relationships is listed in Section 1.4.2 of the Supplementary Material. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the supergroup \citep{RN1425}. As an example, let us consider groups \textit{P2} and \textit{P6}. If we ignore the translations in two directions that both groups share, group \textit{P6} consists of the set of rotations \{$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$\}, in which \textit{P2} \{$0^{\circ}$, $180^{\circ}$\} is contained. \textit{P2} is thus a subgroup of \textit{P6}, and \textit{P6} can be generated by combining \textit{P2} with rotations \{$0^{\circ}$, $120^{\circ}$, $240^{\circ}$\}. Because \textit{P2} is repeated three times in \textit{P6}, \textit{P2} is a subgroup of \textit{P6} with index 3 \citep{RN1425}. The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are subgroups of more complex ones \citep{RN1711}.

@@

The two datasets we present here make it possible to assess the extent to which both behavior and brain responses follow the hierarchy of complexity expressed by the subgroup relationships. Based on previous brain imaging work showing that patterns with more axes of symmetry produce greater activity in visual cortex \citep{RN1170, RN1959, RN1725, keefe_2018}, we hypothesized that more complex groups would produce larger SSVEPs. For the psychophysical data, we hypothesized that more complex groups would lead to shorter symmetry detection thresholds, based on previous data showing that under a fixed presentation time, discriminability increases with the number of symmetry axes in the pattern \citep{wagemans_1991}. Our results confirm both hypotheses, and show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds following these relationships at a level that is far beyond chance. The human visual system thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory.

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \citep{RN1170} and rotation \citep{RN1725}, the visual system represents the hierarchical structure of the 17 wallpaper groups, and thus every composition of the four fundamental symmetries (rotation, reflection, translation, glide reflection) which comprise the set of regular textures. Both SSVEP amplitudes and symmetry detection thresholds preserve the hierarchy of complexity among the wallpaper groups that is captured by the subgroup relationships \citep{RN1711}. For the SSVEP, this remarkable consistency was specific to the odd harmonics of the stimulus frequency that are known to capture the symmetry-specific response \citep{RN1725} and to electrodes in a region-of-interest (ROI) over occipital cortex. When the same analysis was done using the odd harmonics from electrodes over parietal cortex (Supplementary Figure 3.3) or even harmonics from electrodes over occipital cortex (Supplementary Figure 3.4), the data was substantially less consistent with the subgroup relationships (yellow and green lines, Figure 4).

The current data provide a description of the visual system’s response to the complete set of symmetries in the two-dimensional plane. Our design precludes us from independently measure the response to \textit{P1}, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 1), we can conclude that the relationships between P1 and all other groups, where \textit{P1} is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. Previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \citep{RN1253}, they tend to sort them into fewer groups than there actually are (4-12 groups) and it is common for exemplars from different wallpaper groups to be sorted in the same group \citep{RN172}. The more controlled two-interval forced choice approach used in the current behavioral experiment allows us to show that more granular representations of wallpaper groups are measurable in behavior.

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. While it should be noted that the correlation is relatively modest ($R^2 = 0.44$), we note that this may be partly due to the fact that the same individuals did not participate in the two experiments. Future work in which behavioral and brain imaging data are collected from the same participants, will help further establish the connection between the two measurements, and tease apart any additional complexity that may not have been captured by the summary statistics we applied here. It has recently been demonstrated that \textit{W}, a measure of perceptual goodness derived from a holographic model of regularity \citep{van_der_helm_1996}, can predict EEG responses \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for patterns that contain symmetry and other types of regularity. The model was formulated based on dot patterns with symmetry axes centered on a single spatial location. It will be important to determine if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane, and test how well it can explain behavioral and brain responses to wallpapers.

@@

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective and occlusion. Near-regular textures, however, abound in the visual world, and can be approximated as deformed versions of the wallpaper groups \citep{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \citep{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \citep{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \citep{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered \textit{supertextures}, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \citep{RN1764}. Wallpaper groups are visually compelling, and symmetries have been widely used in human artistic expression going back to the Neolithic age \citep{RN1949}. If wallpapers are in fact supertextures, this prevalence may be a direct result of the strategy the human visual system has adopted for texture encoding.

@@

\subsection*{Participants}
Twenty-five participants (11 females, mean age 28.7±13.3) took part in the EEG experiment. Their informed consent was obtained before the experiment under a protocol that was approved by the Institutional Review Board of Stanford University. 11 participants (8 females, mean age 20.73±1.21) took part in the psychophysics experiment. All participants had normal or corrected-to-normal vision. Their informed consent was obtained before the experiment under a protocol that was approved by the University of Essex's Ethics Committee.

>>>>

\documentclass[11pt, twoside]{article}
\usepackage{boisik}
\usepackage[OT1]{fontenc}
\usepackage[affil-it]{authblk}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{multicol}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}

\linenumbers
 \geometry{a4paper, total={170mm,257mm}, left=20mm, top=20mm}

@@

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \citep{RN1425}. The wallpaper groups differ in the number and kind of these transformations \textcolor{ForestGreen}{and we can uniquely refer to different groups using crystallographic notation. In brief, most groups are notated by $PXZ$, where $X\in\{1,2,3,4,6\}$ indicates the highest order of rotational symmetry and $Z\in\{m, g\}$ indicates whether the pattern contains \textit{m}irror or \textit{g}lide symmetry. For example, $P4$ contains 4 fold rotation, while $P2M$ contains 2 fold rotation and a mirror symmetry (see Figure XXX). Two of the groups start with a $C$ rather than a $P$, ($CM$ and $CMM$) which indicates that the symmetries are specified relative to a cell that itself contains repetition. Full details of the naming convention and examples of the wallpaper groups can be found on \href{https://en.wikipedia.org/wiki/Wallpaper_group}{wikipedia}}.

In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \citep{RN1425}. The full list of subgroup relationships is listed in Section 1.4.2 of the Supplementary Material. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the supergroup \citep{RN1425}. As an example, let us consider groups \textit{P2} and \textit{P6}. If we ignore the translations in two directions that both groups share, group \textit{P6} consists of the set of rotations \{$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$\}, in which \textit{P2} \{$0^{\circ}$, $180^{\circ}$\} is contained. \textit{P2} is thus a subgroup of \textit{P6}, and \textit{P6} can be generated by combining \textit{P2} with rotations \{$0^{\circ}$, $120^{\circ}$, $240^{\circ}$\}. Because \textit{P2} is repeated three times in \textit{P6}, \textit{P2} is a subgroup of \textit{P6} with index 3 \citep{RN1425}. The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are subgroups of more complex ones \citep{RN1711}.

@@

The two datasets we present here make it possible to assess the extent to which both behavior and brain responses follow the hierarchy of complexity expressed by the subgroup relationships. Based on previous brain imaging work showing that patterns with more axes of symmetry produce greater activity in visual cortex \citep{Sasaki_2005, Tyler_2005,RN1959, RN1725, keefe_2018}, we hypothesized that more complex groups would produce larger SSVEPs. For the psychophysical data, we hypothesized that more complex groups would lead to shorter symmetry detection thresholds, based on previous data showing that under a fixed presentation time, discriminability increases with the number of symmetry axes in the pattern \citep{wagemans_1991}. Our results confirm both hypotheses, and show that activity in human visual cortex is remarkably consistent with the hierarchical relationships between the wallpaper groups, with SSVEP amplitudes and psychophysical thresholds following these relationships at a level that is far beyond chance. The human visual system thus appears to encode all of the fundamental symmetries using a representational structure that closely approximates the subgroup relationships from group theory.

@@

Here we show that beyond merely responding to the elementary symmetry operations of reflection \citep{Sasaki_2005, Tyler_2005} and rotation \citep{RN1725}, the visual system represents the hierarchical structure of the 17 wallpaper groups, and thus every composition of the four fundamental symmetries (rotation, reflection, translation, glide reflection) which comprise the set of regular textures. Both SSVEP amplitudes and symmetry detection thresholds preserve the hierarchy of complexity among the wallpaper groups that is captured by the subgroup relationships \citep{RN1711}. For the SSVEP, this remarkable consistency was specific to the odd harmonics of the stimulus frequency that are known to capture the symmetry-specific response \citep{RN1725} and to electrodes in a region-of-interest (ROI) over occipital cortex. When the same analysis was done using the odd harmonics from electrodes over parietal cortex (Supplementary Figure 3.3) or even harmonics from electrodes over occipital cortex (Supplementary Figure 3.4), the data was substantially less consistent with the subgroup relationships (yellow and green lines, Figure 4).

The current data provide a description of the visual system’s response to the complete set of symmetries in the two-dimensional plane. Our design precludes us from independently measure the response to \textit{P1}, but because each of the 16 other groups produce non-zero odd harmonic amplitudes (see Figure 1), we can conclude that the relationships between P1 and all other groups, where \textit{P1} is the subgroup, are also preserved by the visual system. The subgroup relationships are in many cases not obvious perceptually, and most participants had no knowledge of group theory. Thus, the visual system’s ability to preserve the subgroup hierarchy does not depend on explicit knowledge of the relationships. \textcolor{ForestGreen}{Previous brain-imaging studies have found evidence of parametric responses with the number of reflection symmetry folds \cite{keefe_2018, Sasaki_2005, makin_2016} and with the order of rotation symmetry \cite{RN1725}. Our study is the first demonstration that the brain encodes symmetry in this parametric fashion across every possible combination of different \textit{symmetry types}, and that this parametric encoding is also reflected in behavior.} Previous behavioral experiments have shown that although naïve observers can distinguish many of the wallpaper groups \citep{RN1253}, they tend to sort them into fewer groups than there actually are (4-12 groups) and it is common for exemplars from different wallpaper groups to be sorted in the same group \citep{RN172}. The more controlled two-interval forced choice approach used in the current behavioral experiment allows us to show that more granular representations of wallpaper groups are measurable in behavior.

\textcolor{ForestGreen}{A large literature exists on the \textit{Sustained Posterior Negativity} (SPN), a characteristic negative-going waveform that is known to reflect responses to symmetry and other forms of regularity and structure \citep{makin_2016}. The SPN scales with the proportion of reflection symmetry in displays that contain a mixture of symmetry and noise \cite{makin_formation_2020, palumbo_scaling_2015}, and both reflection, rotation and translation can produce a measurable SPN \cite{makin_electrophysiological_2013}. It has recently been demonstrated that a holographic model of regularity \citep{van_der_helm_1996}, can predict both SPN amplitude \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for dot patterns that contain symmetry and other types of regularity. The available evidence suggests that the SPN and our SSVEP measurements are two distinct methods of isolating the same symmetry-related brain response: When observed in the time-domain, the symmetry-selective odd-harmonic responses produce similarly sustained waveforms (see Figure 1), odd-harmonic SSVEP responses can be measured for dot patterns similar to those used to measure the SPN \citep{RN1954}, and the one event-related study that has been published on the wallpaper groups also produced SPN-like waveforms \citep{RN1959}. Future work should more firmly establish the connection and determine if the SPN can capture responses similarly precise symmetry responses as the SSVEPs presented here. It would also be worthwhile to ask if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane.}

We observe a reliable correlation between our brain imaging and psychophysical data. This suggests that the two measurements reflect the same underlying symmetry representations in visual cortex. \textcolor{ForestGreen}{It should be noted that the correlation is relatively modest ($R^2 = 0.44$). This may be partly due to the fact that different individuals participated in the two experiments. It may also be related to the fact that participants where not doing a symmetry-related task during the SSVEP experiment, but instead monitored the stimuli for brief, intermittent contrast changes. Previous brain imaging studies have found enhanced reflection symmetry responses when participants performed a symmetry-related task \cite{makin_formation_2020, Sasaki_2005, keefe_2018}. We did not manipulate task during our SSVEP recordings, but it is possible that adding a symmetry-related task to our SSVEP experiment would have produced measurements that reflected subgroup relationships to an even higher extent. We note, however, that our performance is already close to ceiling (see Figure 3). It is possible that adding a symmetry-related task would enhance SSVEP amplitudes across all wallpaper groups without improving the discriminality of individual groups (similar to what was observed by Keefe and his co-authors \cite{keefe_2018}). SPN measurements suggest that task-driven processing may be important for detecting symmetries that have been subject to perspective distortion \cite{makin_conditions_2015} although it should be noted that this effect was much less clear in a subsequent functional MRI study \cite{keefe_2018}. Future work in which behavioral and brain imaging data are collected from the same participants, and behavior is manipulated in the SSVEP task, will help further establish the connection between the two measurements, and elucidate the potential contribution of task-related top-down processing to the current results.}

@@

The 17 wallpaper groups are completely regular, and relatively rare in the visual environment, especially when considering distortions due to perspective \textcolor{ForestGreen}{(see above)} and occlusion. Near-regular textures, however, abound in the visual world, \textcolor{ForestGreen}{and can be modeled} as deformed versions of the wallpaper groups \citep{RN1519}. The correspondence between visual cortex responses and group theory demonstrated here may indicate that the visual system represents visual textures using a similar scheme, with the wallpaper groups serving as anchor points in representational space. This framework resembles norm-based encoding strategies that have been proposed for other stimulus classes, most notably faces \citep{RN435}, and leads to the prediction that adaptation to wallpaper patterns should distort perception of near-regular textures, similar to the aftereffects found for faces \citep{RN1768}. Field biologists have demonstrated that animals respond more strongly to exaggerated versions of a learned stimulus, referred to as “supernormal” stimuli \citep{RN1775}. In the norm-based encoding framework, wallpaper groups can be considered \textit{supertextures}, exaggerated examples of the near-regular textures that surround us. Artists may consciously or unconsciously create supernormal stimuli, to capture the essence of the subject and evoke strong responses in the audience \citep{RN1764}. Wallpaper groups are visually compelling, and symmetries have been widely used in human artistic expression going back to the Neolithic age \citep{RN1949}. If wallpapers are in fact supertextures, this prevalence may be a direct result of the strategy the human visual system has adopted for texture encoding.

@@

\subsection*{Participants}
Twenty-five participants (11 females, mean age 28.7 \textpm\ 3.3) took part in the EEG experiment. Their informed consent was obtained before the experiment under a protocol that was approved by the Institutional Review Board of Stanford University. 11 participants (8 females, mean age 20.73 \textpm\ 1.21) took part in the psychophysics experiment. All participants had normal or corrected-to-normal vision. Their informed consent was obtained before the experiment under a protocol that was approved by the University of Essex's Ethics Committee.
